---
title: "fishHook Tutorial"
author: "Zoran Gajic and Marcin Imielinski"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

```

<style type="text/css">
body, td {
   font-size: 16px;
}
pre code, pre, code {
  font-size: 12px;
  color: rgb(20, 20, 20);
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

## Introduction

The `fishHook` R package enables agile statistical analysis of coding and
non-coding mutational recurrence in cancer through generalized linear modeling (GLM)
of somatic mutation densities and their heterogeneity along the genome. `fishHook`
can be applied to the analysis of any collection of genomic intervals (e.g. genes, enhancers, promoters, genomic tiles) or
complex sets of intervals (e.g. genes sets representing pathways, enhancer sets known to interact with a gene).  The `fishHook`
package is integrated with `GenomicRanges` and `data.table` packages, allowing easy incorporation into bioinformatics workflows
that employ the `R/Bioconductor` ecosystem. 

The core of the fishHook framework is to enable nomination of loci following the correction of known covariates of neutral mutation, e.g. chromatin state, replication timing, and nucleotide context. The goal of fishHook is to identify *cancer drivers*, i.e. loci that are under positive somatic selection and accumulate mutations above "background".  This analysis hinges on the application of a correct null / background model, i.e. one that yields near-uniform Q-Q plots for P value distributions. 

Though we provide pre-computed covariates and a "black box" command-line tool that applies several generic exome and whole genome analyses, 
the key power of `fishHook` lies in its customizability.  This includes the ability to easily incorporate custom covariates and provide a framework for the generation and fitting of bespoke models to nominate loci, e.g. modeling variant- and tumor-type specific background mutational processes. 

For installation instructions, please visit the [FishHook github page](https://github.com/mskilab/fishHook). For background, it may help to have some familiarity with [`data.table`](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html), [`GenomicRanges`](https://www.bioconductor.org/packages/3.7/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.pdf), and [`gUtils`](https://github.com/mskilab/gUtils) packages.

If you use fishHook in your work, please cite: 
[Imielinski, Guo, Meyerson. *Cell.* 2017 Jan 26;168(3):460-472](https://www.sciencedirect.com/science/article/pii/S009286741631741X?via%3Dihub).  

## Driver discovery in cancer whole exomes

We will demonstrate a quick whole exome analysis using public TCGA lung
adenocarcinoma mutation data. Additional packages like `gTrack` and
`rtracklayer` will help with data import and visualization, but are not necessary to run `fishHook`.

```{r message=FALSE, warning=FALSE}
library(fishHook)    
library(gTrack)
library(rtracklayer)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)    
library(magrittr)
library(tidyr)
```


### Read in data

Read in the mutation data and additional tracks that we will use in this analysis.

```{r warning = FALSE}
## mutation calls cached from public GDAC Broad firehose https://bit.ly/2sFxWY6
mutations = dt2gr(fread('http://mskilab.com/fishHook/hg19/luad.maf')) ## using data.table::fread to read maf and gUtils::dt2gr to convert to GRanges

## GENCODE v19 genes these are our "hypotheses"
genes = gr.sub(import('http://mskilab.com/fishHook/hg19/gencode.v19.genes.gtf')) # rtracklayer::import reads gtf and gr.sub replaces chr
genes = genes %Q% (gene_type == 'protein_coding') %Q% (level<3)  # %Q% is a gUtils subsetting operator for GRanges

## protein coding CDS definitions
cds = readRDS(gzcon(file('http://mskilab.com/fishHook/hg19/gencode.v19.cds.rds')))

## bigWig file of fractional coverage of hg19 positions by Agilent exome
## we will use this in combination with cds to define eligible positions
exomecov = import('http://mskilab.com/fishHook/hg19/exome_coverage.bw')

```

Take a peek at our mutations
```{r}
head(mutations[, c('Tumor_Sample_Barcode', 'Variant_Type', 'Variant_Classification', 'Reference_Allele', 'Tumor_Seq_Allele2')])

```
### Instantiate FishHook object from events and eligible territory

First we define the "eligible territory".  This is a key component of all
somatic mutational recurrence analyses, since much of the genome is *not covered*
in sequencing studies. For example, in a whole exome sequencing dataset, less
than 2% of the genome is reliably captured. In a targeted sequencing panel, this
fraction will be even smaller.  Even in whole genome sequencing using Illumina
short reads, only a subset (70%) of the genome is reliably callable ([Li
*Bioinformatics* 2014 Oct 15;30(20):2843–8751](https://academic.oup.com/bioinformatics/article/30/20/2843/2422145)).


Eligible territory coverage will influence the  "denominator" of our recurrence analysis, i.e. the number of positions in each hypothesis interval where a mutation *could have possibly been detected*.  If we do not take eligible territory into account we will mis-estimate the background mutation rate in a given region.

To define eligible territory for this whole exome analysis, we will choose the
portion of cds (protein coding) bases that are captured in at least 95% of whole
exomes, which represents about 24MB of genome.

```{r}
eligible = exomecov %Q% which(score>0.95) # %Q% is a gUtils operator for subsetting on GRanges metadata
eligible = reduce(intersect(eligible, cds, ignore.strand = TRUE)) # we intersect and reduce / collapse our prelim eligible intervals with CDS boundaries

```
We define "events" as nonsynonymous mutations.  In this simple model, we will lump together SNVs (of
different flavors), and indels (of different flavors).  (We discuss more complex
models that subdivide mutation types later in the tutorial).

```{r}
events = mutations %Q% (Variant_Classification != 'Silent') 

```
Now that we have loaded our `hypotheses` (i.e. genes), `events`, and `eligible`, we are ready to
create and analyze a basic `FishHook` object.   Under the hood, object creation triggers 
counting of how many events are in the eligible portion of each hypothesis interval.  We provide the
`idcol` parameter so that each tumor sample (as defined by the
`Tumor_Sample_Barcode` column in the `events` `GRanges`) will provide *at  most
one event* to the counts of each interval.


```{r message = FALSE}
fish = Fish(hypotheses = genes[, 'gene_name'], events = events, eligible = eligible, idcol = 'Tumor_Sample_Barcode')
```

```{r echo=FALSE}
fish
```

### Run basic model without covariates

We can `score` this basic `FishHook` object, i.e. compute p values for every
hypothesis, using a simple glm that models a uniform mutation density along the
genome, i.e. the glm fits only an intercept and applies no covariates (after
correcting for the number of eligible bases in each interval).  

The `$res` field of the `FishHook` contains a `data.table` of scoring results.
`$res` has one row per input hypothesis, with P values, FDRs, effect sizes (fold enrichment
above background), observed and predicted event counts and densities, and additional interval
annotations provided by the user in the `hypotheses` `GRanges`.

```{r eval=FALSE}
fish$score()
head(fish$res[order(p), ])
fish$qqp()
```

```{r echo=FALSE,message=FALSE}
fish$score()
head(fish$res[order(p), ][, p:= as.character(p)]) %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
p = fish$qqp()
p
```
<br>You will notice that this Q-Q plot appears curved and inflated, though its slope
(`lambda`) is reasonably near 1.  The low `alpha` value (MLE of the alpha parameter
                                                     of the Gamma distribution),
suggests that the GLM is detecting additional variance 
in the data that is unmodeled by a pure Poisson regression.  Adding covariates
to the model should improve the quality of the fit.

The top hits in the plot (you can hover over them) include *TP53* but also many
unlikely cancer gene candidates.  Among these are olfactory receptors, which are
located in late replicating regions of the human genome and thus accumulate
neutral mutations more frequently ([Lawrence et al 2013 *Nature* Jul 11;499(7457):214-218](https://www.nature.com/articles/nature12213)).

### Add covariates to FishHook model

To address these issues, we will load in data specifying replication timing,
chromatin state, and nucleotide context.  These are all important determinants
of somatic neutral mutation density. We load these data in as `GRanges` objects using functions in 
`data.table`, `rtracklayer`, and `gUtils` packages (however you are free 
to use any `GRanges` import utility of your choice).

We first load in replication timing data as a `GRanges` then instantiate it as a `Covariate`
Replication timing information is contained in the 
`$score` metadata field of `reptime`.  We instantiate it as a covariate of type "numeric" by
specifying field `score`.

```{r}
## replication timing for NHEK obtained from  https://bit.ly/2sRsXT9 and
## converted to rds via rtracklayer::import
reptimedata = readRDS(gzcon(file('http://mskilab.com/fishHook/hg19/RT_NHEK_Keratinocytes_Int92817591_hg19.rds')))

## instantiate covariate around 'score' field, name the Covariate "replication timing"
reptime = Cov(data = reptimedata, field = 'score', name = 'ReplicationTiming') 

```

Below, `context` is a `GRanges` object with 98 columns representing tri, di, and mononucleotide context counts in the
hg19 genome.  Code for computing context (e.g. for another genome) is provided [*here*](http://lmgtfy.com/?q=under+construction).

We instantiate a numeric `Covariate` object from `context`, choosing only two of the
columns here to take into account G and C content.  Note that the covariate object
can be vectorized (concatenated, subsetted) and instantiated around several
columns of an input `GRanges`.  As a result, `contextcov` will be length 2
(representing G and C nucleotide fraction).

```{r}
context = readRDS(gzcon(file('http://mskilab.com/fishHook/hg19/nucleotide.context.rds')))
gc = Cov(data = context, field = c('C', 'G')) ## instantiate Covariate around G and C fields

```

Finally we load in chromHMM data for cell line A549 from Epigenomics Roadmap.
We will want to create a covariate that will model the fraction of
heterochromatic and quiescent regions in each query interval.

To do so, we will create an "interval" covariate by not specifying a metadata
field.  

```{r}
### data cached from https://egg2.wustl.edu/roadmap/data/byFileType/chromhmmSegmentations/ChmmModels/coreMarks/jointModel/final/E114_15_coreMarks_mnemonics.bed.gz
chromhmm = gr.sub(import('http://mskilab.com/fishHook/hg19/E114_15_coreMarks_mnemonics.bed.gz'), 'chr', '') ## import from bed then gUtils::gr.sub to strip 'chr' identifier
hetchromdata = chromhmm %Q% (name %in% c('8_ZNF/Rpts', '9_Het', '15_Quies')) # %Q% is gUtils operator for subsetting on GRanges metadata, in this case selecting for heterochromatic regions
hetchrom = Cov(hetchromdata, name = 'Heterochromatin') ## instantiate interval covariate
```

We now add these covariates to the model. For type `numeric` covariates, e.g. replication
timing, this will trigger the calculation of the average value of each
covariate within the eligible subset of each hypothesis interval. and the fractional overlap of of its eligible subset
This annotation is the most computationally intensive and slowest aspect of
fishHook analyses, though occurs within a few seconds for this small number of covariates.

```{r message=TRUE}
## note how we can concatenate Covariate objects with c() operator
fish$covariates = c(reptime, gc, hetchrom)

```

Now that we've added covariates, we can re-score `fish` and compute p values.  Looking at these
results, we immediately see an improvement in `lambda` (closer to one) and `alpha`
(increased), and the nominated gene list (no more olfactory receptors, We
see also a reasonable number of significant (fdr<0.1) genes
at the top of the list (or at the top right of the QQ plot) that have been
biologically implicated in lung adenocarcinoma tumorigenesis.


```{r eval=FALSE}
fish$score()
fish$res[fdr<0.25, ][order(p), ]
fish$qqp()
```
```{r echo=FALSE,message=FALSE}
fish$score()
fish$res[order(p), ][fdr<0.25, ][, p:= as.character(p)] %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
p = fish$qqp()
p
```
<br>
Note that this model is still quite rudimentary (e.g. we have lumped together
 SNVs and indels, we have not substratified SNVs by mutational context, we have
employed very few covariates) but we still get a reasonable gene list and minimal
statistical inflation. 

### Merge additional covariates

We may notice that some of the significantly mutated genes are not known to 
be expressed in lung adenocarcinoma tissues.  We also know that genes that 
are more highly expressed in a given tissue have lower mutation rates.  We want to merge
in gene expression into the model, but not start from scratch (i.e. create a new model).
We can do this using the `$merge` function. 


```{r}
## load GRanges of lung adenocarcinoma average gene experssion
exprdata = readRDS(gzcon(file('http://mskilab.com/fishHook/hg19/lung.expr.rds'))) 

# log transform expression values
exprdata$log.tpm = log10(exprdata$tpm+0.01) 

## create Covariate for log gene expression, using log.tpm field in the exprdata object
expr = Cov(exprdata, field = 'log.tpm', name = 'LungExpression') 

## merge / append this new covariate into the FishHook object
fish$merge(expr) 

## look at the $data field to see the new merged covariate data as an additional column
head(fish$data)
```

### Subset the FishHook object

Now that we've merged gene expression into this model, we would like to apply it both as a covariate and as a gene filter.
We would like to exclude from the analysis genes that are known to have poor expression in lung adenocarcinoma tissue, because these
genes are unlikely to harbor driver alterations. 

The dimension of the fish object is hypotheses by covariates.  
To subset rows (i.e. hypotheses) from this model we can use the subsetting feature of the FishHook object via the `[` operator.  
We will use this functionality to apply a strict "expression filter" and keep only genes that are expressed >10 TPM in lung adenocarcinoma.
We will then re-score this subsetted model.  

```{r}
fish = fish[which(fish$data$LungExpression>1), ] ## subset for high lung expression
```

```{r echo=FALSE}
fish
```
This object now contains 8765 hypotheses (i.e. genes), which we can re-score to obtain P values and FDRs. 


```{r eval=FALSE}
fish$score()
fish$res[fdr<0.25, ][order(p), ]
fish$qqp()
```
```{r echo=FALSE,message=FALSE}
fish$score()
fish$res[order(p), ][fdr<0.25, ][, p:= as.character(p)] %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
p = fish$qqp()
p
```

<br>
To inspect the parameters of this model and see which features it is using, we
can employ the`$model`accessor:

```{r size="tiny"}
summary(fish$model)
```

We can see from the `Estimate` and `Pr(|>z|)` columns of the `Coefficients` table that replication timing and lung gene expression are significantly
negatively correlated and heterochromatin is significantly positively correlated
with mutational density (as expected).  However, this table shows that `G` and `C` content are not significantly correlated with 
mutation density.  We can use the column subsetting function to remove these covariates and re-score to see how the results change. 


```{r}
fish2 = fish[, -c(2:3)] ## remove the 2nd and 3rd covariates ('G', 'C')
fish2$score() ## re-score

## check if our new top genes are identical to the previous - they are
identical(fish2$res[fdr<0.25, sort(gene_name)], fish$res[fdr<0.25, sort(gene_name)])
```
The results are identical with and without `G` and `C`  covariates.  This further suggests that these covariates are not necessary in the model and can likely be excluded.


### Analyze Reactome pathways

Read in and parse reactome pathways into list of gene symbols, then match against genes in our model.

```{r}
## parse Reactome pathways from .gmt format
pathways = strsplit(readLines('http://mskilab.com/fishHook/hg19/ReactomePathways.gmt'), '\t')
pathways = structure(lapply(pathways, '[', -1), names = sapply(pathways, '[', 1))

## match them to create sets of indices as a named list
sets = sapply(sapply(pathways, match, fish$hypotheses$gene_name), setdiff, NA)
```

Here is what the pathways and sets look like:


```{r}
head(pathways[1:2])

head(sets[1:2])
```
<br>
The list `sets` contains integer vectors that index `fish$hypotheses`.

To run a set analysis, we just set the `$sets` variable in the `FishHook` object.
This triggers scoring of hypothesis *sets* (in this case gene
sets). The results of the set analysis are shown in $setres variable. 

The set analysis is a bit more computationally intensive.  We can speed things up
through parallelization (setting `fish$mc.cores = 5).


```{r eval=FALSE,echo=TRUE}
fish$mc.cores = 5

## this triggers scoring of gene sets using covariate corrected model
fish$sets = sets

## list top sets
fish$setres[order(p)[1:5], ]
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
fish$mc.cores = 50
fish$sets = sets
head(fish$setres[order(p), ][1:5, ][, p:= as.character(p)]) %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")

```
<br>
Examining the results table, we can see that most of the significant pathways appear
related to *TP53*. Indeed, if we inspect the hypotheses (i.e. genes) contributing to
these these gene sets, we will see that they are dominated by *TP53* and 1-2
additional genes. For example:

```{r eval=FALSE}
## pick a set 
mysetname = fish$setres[order(p)[1], setname]

## use sets list to index fish results, sorting by p value
fish$res[fish$sets[[mysetname]], ][order(p), ]
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
fish$res[sets[[fish$setres[order(p), ][fdr<0.1,  ][1, setname]]], ][order(p),
][, p:= as.character(p)] %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
```

<br>
This is a common challenge with pathway analysis of mutations, since many cancer pathways are usually driven by a single "celebrity" gene. We can dig a little deeper to identify significant gene sets that do not have
*TP53* . Here is one approach:

```{r eval = FALSE}
## these are gene sets with TP53
has.tp53 = sapply(fish$sets, function(x) any(fish$hypotheses$gene_name[x] %in% 'TP53'))

## we subset on gene sets that do not have TP53
head(fish$setres[!has.tp53, ][order(p), ])
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
has.tp53 = sapply(fish$sets, function(x) any(fish$hypotheses$gene_name[x] %in% 'TP53'))

head(fish$setres[!has.tp53, ][order(p), ][, p:= as.character(p)]) %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
```

<br>
These sets appear interesting and are related to  additional biological processes known 
to be important to lung adenocarcinoma biology, such EGFR signaling. The top gene set "Extracellular matrix organization" appears especially interesting, because it is not obviously associated with known targets of driver mutations in this disease. Let's examine genes contributing to this top gene set:

```{r eval = FALSE}
## take the top non-TP53 associated gene set 
mysetname = fish$setres[!has.tp53, ][order(p)[1], setname]

## use sets list to pull the individual hypothesis results for members of this set
head(fish$res[fish$sets[[mysetname]], ][order(p), ])
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
## take the top non-TP53 associated gene set (i.e. "Extracellular matrix organization")
mysetname = fish$setres[!has.tp53, ][order(p)[1], setname]

## use sets list to pull the individual hypothesis results for members of this set
head(fish$res[fish$sets[[mysetname]], ][order(p), ][, p:= as.character(p)])  %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")

```
<br>
Interesting! This significant gene set is composed of extracellular matrix genes
(*COL5A2*, *ITGAL*) that are not significant on their own.  This constitutes a
true pathway level hit. 


### Analyze truncating mutations 

We can run a similar analysis but choosing only truncating mutations (by
subsetting the mutation GRanges). Scoring this new model, we obtain a different mutation list, 
containing likely candidate drivers with enrichment of frameshift, nonsense, or
nonstop indels and SNVs.  

```{r eval=FALSE}
## replace events with new subset of mutations (using %Q% subsetting operator from gUtils)
fish$events = mutations %Q% 
   (grepl('(Frame_Shift_)|(Nonsense)|(OutOfFrame)|(Nonstop)', Variant_Classification))

## re-score model and inspect results
fish$score() 
fish$res[order(p), ][fdr<0.25, ]
fish$qqp()
```

```{r echo=FALSE,message=FALSE,warning=FALSE}
fish$events = mutations %Q% (grepl('(Frame_Shift_)|(Nonsense)|(OutOfFrame)|(Nonstop)', Variant_Classification))
fish$score()
fish$res[order(p), ][fdr<0.25, ][, p:= as.character(p)] %>% kable() %>%  kable_styling(bootstrap_options =
c("striped", "hover", "condensed", "responsive")) %>%
    scroll_box(width = "100%", height = "200px")
p = fish$qqp()
p
```

Though *TP53*, *RBM10*, *SETD2*, and *ARID1A* are well known targets of truncating mutations in lung adenocarcinoma, 
[*IL32*](https://www.genecards.org/cgi-bin/carddisp.pl?gene=IL32), [*TRIB1*](https://www.genecards.org/cgi-bin/carddisp.pl?gene=IL32), [*SUN1*](https://www.genecards.org/cgi-bin/carddisp.pl?gene=IL32) are interesting candidates that have not previously
been associated with lung adenocarcinoma. 
